Measuring the double-drift illusion and its resets with hand trajectories

### 2022-01-18 Manuscript.Rmd
revisions from Patrick

---
title: "Measuring the double-drift illusion and its resets with hand trajectories"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document:
    reference_docx: template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

_1: Centre for Vision Research, York University, Toronto, ON M3J1P3, Canada_\
_2: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\
_3: Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NJ 03755, USA_\

@@

The research was support by funding from NSERC Canada (PC and DYPH), as well as by the Department of Psychological and Brain Science, Dartmouth College (PC).

@@

We compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller et al. [-@Heller2021]. A more sophisticated model allows predicting perceived speed as well [@Johnston2013], but this model predicts perceived direction and speed for a mixture of first order motions, and it is unclear if this works for a combination of first and second order motion (what we call internal and external speed here). In the vector average model we use here, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$. That is, the deviation from the physical direction is given by:

@@

The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it? In Figure 2D, we compare mean illusion strength to that reported previously [@Heller2021]. As can be seen the averages from our data here (red circles) coincide fairly well with the illusion strength predicted by the model fitted to all their data (Eq. 1, black line). If we fit the vector combination model, we get $K=0.738$ which is very close to $K=0.74$ reported by Heller et al. [-@Heller2021]. That is; tracking the illusion while observing it does not change the strength of the illusion.

@@

The X coordinates of the detected reset points are used as the spatial offset of the reset. We also need the time of the reset. Using the Y coordinates of the drawn trajectory (divided by speed) would assume that the real and perceived vertical position of the gabor are the same, which would imply that for very strong illusions the perceived speed can be many times the real speed. This seems unrealistic. As described above, it is unclear if the overall perceived speed of a mixture of many first order motions [@Johnston2013] applies to the double-drift illusion and if so, whether or not that can be used to calculate perceived time from path length. Instead, we chose a simplification by using the radius (R; distance from origin to reset, divided by speed) as the reset time. However, while useful for the current project, it remains to be tested how perceived speed and perceived path length are related to each other and to stimulus properties in the double-drift illusion.

To facilitate modeling, the horizontal and vertical (straight ahead) coordinates of the reset points are kept isometric by giving them in centimeters. The distance from origin can then be converted to time as 1 _cm_ = 4/13.5 _s_ in the 4 _s_ condition and 1 _cm_ = 3/13.5 _s_ in the 3 _s_ condition. For each model using a probability density function to estimate model likelihood, the variable that is used is divided by its median value such that each of their probability densities are on a comparable scale.

@@

```{r fig5, echo=FALSE, message=FALSE, fig.width=6, fig.height=5.333333, fig.cap="\\label{fig:fig5}**Figure 5:** Reset models. **A:** Model limits relative to reset point locations. Reset points neither fall close to a spatial limit (blue line) or a temporal limit (dark red line: 4 _s_ passes, light red line: 3 _s_ passes). Marginal distributions show that both the time and spatial offset of reset points approximate a gamma distribution. **B,C:** Color illustrates the gamma probability distributions, where white is zero probability, and the darkest red or blue is the maximum probability density. This is a relatively scale-free unit, where the total surface of the full probability density function (along a line orthogonal to the predictor variable) is 1. **B:** Gamma distribution on time (log likelihood: -605.6). Dark and light red lines: denote the median reset time for 4 _s_ and 3 _s_ passes in angular bins with an equal number of resets. **C:** Gamma distribution on spatial offset (log likelihood: -632.6). Blue lines indicate median reset offsets in Y-coordinate bins with an equal number of resets."}
invisible(capture.output(plotModels()))
```

For the space and time limit with normal distributions (means in Fig 5A, blue: spatial, red: time), the time-based distribution fits better for the group data (normal space AIC=2,834.83, normal time _AIC_=1,677.98; _p_<.001) as well as all individual participants, and the same holds within gamma distribution models of reset models (gamma space AIC=1,269.18, gamma time AIC=1,215.12; _p_<.001). Furthermore, the gamma-based distribution on time of resets performed better than the normal distribution of reset time in the pooled data ( _p_<.001 ), and this holds for most participants (6/9) as well. This means that time-based distributions explain our reset points better than models assuming a spatial offset, so that the time-based gamma distribution was the most successful of all 4.

@@

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in their experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the distribution of resets throughout time are largely in line with a mechanism based on random distractions of attention, which may involve eye-movement planning and execution. The rate of resets that we found here, ~0.75 / _s_, is too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements. On the other hand, gamma distributions [e.g. @McGill1965; @Wolfe2002] do model a random occurrence of events (rate parameter) where only some fraction of these trigger an action (shape parameter). In other words, the good fit of a gamma distribution may reveal some properties of the process underlying resets of the double-drift illusion. While the process giving rise to resets is likely memoryless, a process giving rise to a gamma distribution (with shape > 1) is not, it does capture two ideas: 1) resets are triggered by random events (such as distractions of attention or microsaccades), and 2) only some of these actually lead to a reset. An ex-Gaussian distribution does not fit the data as well as a gamma distribution, but could be a more realistic alternative. However, which process actually leads to resets remains to be seen in future work.

@@

The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, green dots). This can be explained if spontaneous resets occurred randomly over time, so that longer durations would have more resets and a smaller final offset at the endpoint. Liu et al. [-@Liu2021] modeled this with a Poisson process having a constant chance of reset of ~0.7 / _s_ (Fig 6B, orange line). In comparison, in our experiment, the gamma distribution fit to the data of explicit resets (rate=$2.81$, shape=$3.74$), gave a ~0.75 / _s_ chance of a reset, a remarkaly similar estimate. That's why, we also used the parameters of our gamma distribution fit from data to generate the PSEs that should be found in Liu et al.'s experiment with no additional fitting (Fig 6B, dashed orange line). We see that the two findings, based on very different approaches, are mostly in agreement with each other.

>>>>

---
title: "Measuring the double-drift illusion and its resets with hand trajectories"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document:
    reference_docx: template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

_1: Centre for Vision Research, York University, Toronto, ON M3J1P3, Canada_\
_2: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\

@@

The research was support by funding from NSERC Canada (PC and DYPH).

@@

We compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller et al. [-@Heller2021]. A more sophisticated model allows predicting perceived speed as well [@Johnston2013], but this model predicts perceived direction and speed for a mixture of first order motions, and it is unclear if this works for a combination of first and second order motion (what we call internal and external speed here). In the vector average model we use here, the perceived direction

To determine if hand tracking produces an illusion similar to that seen in other studies, we compare our results to those reported by Guernsey and Biard [-@Guernsey2012], Cavanagh and Tse [-@Cavanagh2019] and Heller et al. [-@Heller2021] who all used similar displays. We also compare the illusory direction of the heading against the predictions of a simple vector combination model [@Tse2006; @Cavanagh2019; @Heller2021]. There are alternative models that could be considered, one from Kwon et al. [-@Kwon2015] and another from Johnston and Scarfe [-@Johnston2013]. Kwon et al.’s [-@Kwon2015] Kalman filter tracking model assumes a saturation of the position shifts by 200 ms, something we do not observe here. Johnson and Scarfe’s [-@Johnston2013] model used a harmonic average to combine vectors, a process that gives more weight to vectors with lower magnitudes. This model was used to estimate the perceived direction of an array of stationary Gabors with various internal motions but is less suited to the double-drift stimulus that has only two vectors, one of which is a second-order motion (the motion of the Gabor itself). In the simple vector combination model we use here, the perceived direction is a vector combination of the external ($V_e$) and internal ($V_i$) motions weighted by a constant $K$. That is, the deviation from the physical direction is given by:

@@

The illusion size we find with tracking appears to be in the range of that reported in other article for similar speeds and eccentricity. For example, Heller et al. (2021) reported an illusion of 30° for internal and external speeds of 7.2 and 7.2 dva/s compared to the 27.8° here (average of left and right 3Hz internal motion) with internal and external speeds both 7 dva/s. The data clearly show the illusion, and we can compare the mean illusion strengths here to those of Heller et al. [-@Heller2021] in more detail using their vector combination model. The averages from our data here (Fig 2D: red circles) do coincide fairly well with the illusion strength predicted by the model fitted to their data. The best fit value of the vector combination model (Eq. 1 above; Fig 2D: black line) to our data $K=0.738$ is very close to $K=0.74$ reported by Heller et al. [-@Heller2021]. We can conclude that tracking the illusion while observing it does not appear to change its strength.

@@

The X coordinates of the detected reset points give the deviation from the vertical path and are used as the spatial offset of the reset. We also need the time of the reset. The gabor moves at a uniform speed along the Y axis so distance along Y denotes the elapsed time for the physical path. However, the illusory path of the gabor does not have to stay in alignment with its physical location on the Y axis. If it did, its speed and path length would have to increase for larger illusion angles: for example, 41% faster and farther for a 45° illusion, 73% faster and farther for a 60° illusion. Observers have not reported these increases in speed or path length compared to the control with no internal motion. Moreover, in a recent unpublished study we directly measured the perceived path length of the illusion. The path length was always fixed proportion of the physical path (around 70%) for all illusion strengths, Gabor speeds and physical path lengths.  This means that the perceived speed must be a fixed proportion of the physical speed as well and therefore, in the tracings, it is the distance along the traced path (the radius in our diagrams) that tells us the time since the trial began — not the distance along the physical path (the Y axis). For completeness, we also analyze the distributions along the Y axis.

To facilitate modeling, the horizontal (illusory deviation) and vertical (straight ahead) coordinates of the reset points are kept isometric by giving them in centimeters. The distance from origin can then be converted to time as 1 _cm_ = 4/13.5 _s_ in the 4 _s_ condition and 1 _cm_ = 3/13.5 _s_ in the 3 _s_ condition. A probability density function is used to estimate each model’s likelihood and the variable that is used (X or T) is divided by its median value so that each of their probability densities are on a comparable scale.

@@

```{r fig5, echo=FALSE, message=FALSE, fig.width=6, fig.height=5.333333, fig.cap="\\label{fig:fig5}**Figure 5:** Reset models. **A:** Model limits relative to reset point locations. Reset points neither fall close to a spatial limit (blue line) or a temporal limit (dark red line: 4 _s_ passes, light red line: 3 _s_ passes). Marginal distributions show that both the time and spatial offset of reset points approximate a gamma distribution. **B,C:** Color illustrates the gamma probability distributions, where white is zero probability, and the darkest red or blue is the maximum probability density. **B:** Gamma distribution on time (_AIC_ = 1215.12). Dark and light red lines: denote the median reset time for 4 _s_ and 3 _s_ passes in angular bins with an equal number of resets. **C:** Gamma distribution on spatial offset (_AIC_ = 1269.18). Blue lines indicate median reset offsets in Y-coordinate bins with an equal number of resets."}
invisible(capture.output(plotModels()))
```

For the space and time limit with normal distributions (means in Fig 5A, blue: spatial, red: time), the time-based distribution fits better for the group data (normal space _AIC_=2,834.83, normal time _AIC_=1,677.98; _p_<.001) as well as all individual participants, and the same holds within gamma distribution models of reset models (gamma space _AIC_=1,269.18, gamma time _AIC_=1,215.12; _p_<.001). Furthermore, the gamma-based distribution on time of resets performed better than the normal distribution of reset time in the pooled data ( _p_<.001 ), and this holds for most participants (6/9) as well. This means that time-based distributions explain our reset points better than models assuming a spatial offset, so that the time-based gamma distribution was the most successful of all 4.

@@

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in their experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the distribution of resets throughout time are largely in line with a mechanism based on random distractions of attention, which may involve eye-movement planning and execution. The rate of resets that we found here, ~0.75 / _s_, is too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements. On the other hand, gamma distributions [e.g. @McGill1965; @Wolfe2002] do model a random occurrence of events (rate parameter) where only some fraction of these trigger an action (shape parameter). In other words, the good fit of a gamma distribution may reveal some properties of the process underlying resets of the double-drift illusion. While the process giving rise to resets is likely memoryless, a process giving rise to a gamma distribution (with shape > 1) is not, it does capture two ideas: 1) resets are triggered by random events (such as distractions of attention or microsaccades), and 2) only some of these actually lead to a reset. An alternative to the gamma distribution, the ex-Gaussian (ref), could be a more realistic alternative but does not fit the data as well. However, the process that actually leads to resets remains to be seen in future work.

@@

The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, green dots). This can be explained if spontaneous resets occurred randomly over time, so that longer durations would have more resets and a smaller final offset at the endpoint. Liu et al. [-@Liu2021] modeled this with a Poisson process having a constant chance of reset of ~0.7 / _s_ (Fig 6B, orange line). In comparison, in our experiment, the gamma distribution fit to the data of explicit resets (rate=$2.81$, shape=$3.74$), gave a ~0.75 / _s_ chance of a reset, a remarkably similar estimate. We then used the parameters of our gamma distribution fit to our data to generate the PSEs that should be found in Liu et al.'s experiment with no additional fitting (Fig 6B, dashed orange line). We see that the two findings, based on very different approaches, are mostly in agreement with each other.
